---
title: "03 Indexing Datasets"
author: "Bell Wu"
date: '2022-06-07'
output: html_document
---

Today we will subset data using R...
  
###### Subsetting Data
  
- most basic way to subset data  
`example[x, y]`
- example is dataset, x = row, y = columns  
  
Following an [example](https://www.r-bloggers.com/2016/11/5-ways-to-subset-a-data-frame-in-r/) to subset: 
```{r}
# import online dataset
education <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/robustbase/education.csv", stringsAsFactors = FALSE)
colnames(education) <- c("X","State","Region","Urban.Population","Per.Capita.Income","Minor.Population","Education.Expenditures")
View(education)
```  
Task:
- Want Midwestern division of schools  
- calculate money spent per child for each state in region
  
Variables required:  
- State, Minor.population, Education.expenditures  
- Only need rows that correspond to variable 2  
  
Method 1 (using square brackets):

```{r}
ed_exp1 <- education[c(10:21), c(2, 6:7)] # use to subset what you want
ed_exp2 <- education[-c(1:9, 22:50), -c(1, 3:5)] # use to remove what you don't want
```
- this method becomes tedious if large dataset and don't know index values of rows/columns
  
Method 2 (using "which" function):  
  
```{r}
ed_exp3 <- education[which(education$Region == 2), names(education) %in% c("State", 
                                                                           "Minor.Population",
                                                                           "Education.Expenditure")]
```

Code explanation:  
- which function is used to extract the rows:
  - this function returns indices where Region column of education data from is 2
```{r}
ed_exp3_which <- education[which(education$Region == 2), ] # blank after column = return all columns
View((ed_exp3_which))
```
  - as result will give us the rows we need
  - so to retrieve the columns we need use "%in%" operator on names of education data frame
  
Method 3 (using "subset" function):  

```{r}
ed_exp4 <- subset(education, Region == 2, select = c("State", "Minor.Population", "Education.Expenditure"))
```
  
Code explanation:  
- subset function takes 3 arguments: 
  - data frame want subsetted
  - rows corresponding to condition which you want it subsetted (subset where region =2)
  - columns you want returned (select state, minor.population, education.expendature as columns)
  
Method 4 (using "dplyr"):  
  
```{r}
install.packages("dplyr")
library(dplyr)
ed_exp5 <- select(filter(education, Region == 2), c(State, Minor.Population:Education.Expenditures))
```
- involves using dplyr package
- filter:
  - first argument is data frame
  - second argument is condition by which we want subsetted 
  - result is entire data frame with only rows we wanted
- select:
  - first argument is data frame
  - second argument is names of columns we want selected (don't need names function and quotations)

  
  
  
  
.







