---
title: "04 For loops in R"
author: "Bell Wu"
date: '2022-06-07'
output: html_document
---
  
#### For loops  
  
For loops are important as they are used to iterate over a vector  
  
##### Syntax of for loop:
> for (val in sequence)  
> {  
> statement  
> }  

- `sequence` is vector
- `val` takes on each of its value during loop
- `statement` is evaluated in each iteration  
  
Example (how for loops work):  
```{r}
for (x in 1:10){
  print(x)
}
```
- note how each individual number of vector from 1-10 is printed  
- for works  like an iterator method 
  
(cont'd):
```{r}
fruits <- list("apple", "banana", "cherry")
for (x in fruits){
  print(x)
}
```  
- this can also be done using characters 
*note that for does not require indexing variable to be set beforehand (unlike while loops)*
  
Example (adding `Break` into statement):
```{r}
fruits <- list("apple", "banana", "cherry", "orange")

for (x in fruits) {
  if (x == "cherry") {
    break
  }
  print(x)
}
```
- note that loop stops printing at cherry since we chose to finish the loop (via break) when x is cherry 
  
  
Example (count number of even numbers in vector):  
```{r}
x <- sample.int(50, 10)
count <- 0
for (val in x){
  if(val %% 2 == 0)
    count = count + 1
}
print(count)
```




