---
title: "05 Generating colour palettes"
author: "Bell Wu"
date: '2022-06-09'
output: html_document
---
##### [Generating colour palettes:](https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html)
  
###### brewer.pal()  
Usage: `brewer.pal(n, name)`  
- n = number of differnt colours in palette, min 3 and max depending on palette
- name = a palette name from lists  
Details:  
- `display.brewer.pal(n, name)`  
- displays selected palette in graphics window  
- `display.brewer.all()`:  
- displays few palettes simultaneously in graphics window 
- `brewer.pal.info`:  
- returns information about available palettes as data frame; this is a variable  
  
###### colorRampPalette:
Usage: `colorRampPalette(palett, bias = 1, method = c("spline", "linear"))`  
- palette = a vector of colours  
- bias = bias > 1 puts more colours at high values
- method = interpolation method. 
*Explanation and use of function:*  
- colorRamp()
```{r}
pal <- colorRamp(c("red", "blue"))
pal(0)

pal(0.5) # can pass any value between 0 and 1 to pal() function

pal(seq(0, 1, len = 10)) # can also pass sequence of numbers to pal() function
```
- pal is function returned by colorRamp()
- pal(0) returns 1x3 matrix: range 0-255 indicate quantities of red in RGB columns 1, 2, 3
- colorRamp() gives function that can interpolate between multiple colours
  
- colorRampPalette()
```{r}
pal <- colorRampPalette(c("red", "yellow"))
pal(2) # just returns two colors: yellow and red


```
- similar to colorRamp but gives FIXED number of colours that interpolate palette
- colours returned represented by hexadecimal strings

###### RColorBrewer package  
Package components:  
- offers three types of palettes: 
  - sequential: ordered numerical data
  - diverging: numerical data that can be positive or negative, represnt deviatiosn from some norm or baseline
  - qualitative: for qualitative unordered data







