---
title: "07 George Dataset Subtyping"
author: "Bell Wu"
date: '2022-06-06'
output: html_document
---

First need to write function to pull genes
```{r}
setwd("/Users/benson/R_programming/R_hmap_test")
George <- read.csv("George_nodup.csv")
head(George)
rownames(George) <- George$X
George <- George[ , -1]
```

##### important functions:

###### max.col:
- finds maximum position for *each row* of a mtarix, breaking ties at random
`max.col(m, ties.method = c("random", "first", "last"))`
  - m = numerical matrix
  - ties.method = character string specifying how ties are handled, "random" by default
  - returns index of maximum value for each row, an integer vector of length nrow(m)
  
###### max and min:
- returns (regular or parallel) maxima and minima of input values  
`max(..., na.rm = FALSE)`  
  - ... = numeric or character arguments  
  - na.rm = logical indicating whether missing values should be removed  
*Details*:  
- return max or min of all values present in arguments:
  - as integer (logical or integer)
  - as double if all are numeric  
`pmax and pmin`:  
- take one or more vectors (or matrices) as arguments
- return single vector giving "parallel" maxima of vectors
Example:  
```{r}
testG <- George[ , c(1:3)]
markerGene = c("ASCL1", "NEUROD1", "YAP1", "POU2F3")
testG_m <- subset(testG, subset = rownames(testG) %in% markerGene )
pmax(testG_m[ , 1], testG_m[ , 2])
# output: 975.988000   4.861290   0.128793   1.495010
```
- returns the max between the rows of matrices (or columns)  
  
###### which.max():  
- determines location (index) of first min or max of numerical or logical vector 
*SYNTAX:* `which.max(dataframe_name$columnname)`  
  
###### rownames(x)[i]  
- x = dataframe, i = number for index
- will give the rowname from dataframe based on the integer value used to index
  
Now to write the function...  
##### Function for pulling highest gene count
```{r}

mGenes = c("ASCL1", "NEUROD1", "YAP1", "POU2F3")

subtypeData <- function(counts, markerGene) { # create function for counts and genes of interest
  mcounts <- subset(counts, subset = rownames(counts) %in% markerGene ) # make matrix of markers of interest
  r_index <- sapply(mcounts, which.max) # use sapply to find highest max value from marker genes
  hgenes <- rownames(mcounts)[r_index] # use rownames to index the name of gene that is highest
  scounts <- rbind.data.frame(hgenes, counts) # generate dataframe by binding the rows from each "matrix"
  return(scounts)
}

sGeorge <- subtypeData(counts = George, markerGene = mGenes)
```
  
##### Identify subtype distribution in George dataset
```{r}
## to count how many subtypes in this dataset
subtypes <- sGeorge[1, ]
rownames(subtypes) = "subtype"

occurences <- table(unlist(subtypes)) # unlist dataframe to character vector, use table() to count occurences
as.matrix(occurences)

# plot a pie chart from the different subtypes
slices <- occurences
lbls <- rownames(occurences)
percent <- round(slices/sum(slices) * 100)
lbls <- paste(lbls, percent)
lbls <- paste(lbls, "%", sep = " ") # add % to the labels

library(RColorBrewer)
pcolour <- colorRampPalette(brewer.pal(4, "Set1"))(4)

George.pie <- pie(slices, labels = lbls, col = pcolour, main = "Subtype Distribution of George dataset")

setwd("/Users/benson/R_programming")
dev.new()
pdf("George_subtype_distribution(2).pdf", width = 4, height = 4)
pie(slices, labels = lbls, col = pcolour, main = "Subtype Distribution of George dataset")
dev.off()
```
##### Helpful functions:

###### names():
- Functions to get or set the names of an object

  
###### dplyr::count()
- quickly count unique values of one or more variables
`df %>% count (a,b)`

###### table():  
```
table(...,
      exclude = if (useNA == "no") c(NA, NaN),
      useNA = c("no", "ifany", "always"),
      dnn = list.names(...), deparse.level = 1)
```
- builds contingency table of counts at each combination of fators  
- ... = one or more objects which can be interpreted as factors, or list whose components can be interpreted  
  
###### paste()  
- concatenante vectors after converting to character
`paste(..., sep = " ", collapse = NULL, recycle0 = FALSE)`
- ... = one or more R objects, to be converted to character vectors
- sep = character string to separate terms
- collapse = optional character string to separate the results
  
  
Now to generate a heatmap from this:
  
need libraries:
- [pheatmap](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12/topics/pheatmap)
  
###### To test out scaling functions when generating heatmap:
```{r}
htestG <- George[c(1:10), c(1:3)] # take only first 10 rows since code runs long if not
# pheatmap(htestG) # testing pheatmap function

# also want to test out scaling factors to calculate zscores 
cal_z_score_1 <- function(x){ # plot in blue
(x - mean(x)) / sd(x)
}

cal_z_score_2 <- function(x){ # plot in red
  t(scale(t(log2(1+x))))
}

z1_testG <- t(apply(htestG, 1, cal_z_score_1))
z2_testG <- cal_z_score_2(htestG)
# now have the scaled data, plot density plots to see distribution
d1 <- density(z1_testG[ , 1]) # returns density data
d2 <- density(z2_testG[ , 1])
plot(d1, xlab = "z-scaled counts of patient S00022", # plot density of plot 1st function "d1"
     ylab = "Density",
     main = "Comparison between scaling functions and raw data",
     col = "red") 

lines(d2, col = "blue") # lines will add to exiting plot but not make new one

legend("topright", legend = c("scaled 1", "scaled 2"), 
       col = c("red", "blue"),
       lty = "solid")
# blue distribution looks cleaner; to test out on heatmap to see scaling
```
  
###### Now to test scaling function on heatmap distribution
```{r}
hcolour <- colorRampPalette(brewer.pal(8, "RdBu"))(100) # generate colour panel

test_count <- sGeorge
test_count <- as.matrix(count[-1, ])
mode(test_count) = "numeric"
geneSet <- c("CD63", "CD163")

cal_z_score_1 <- function(x){ # plot in blue
(x - mean(x)) / sd(x)
}

#scaling with cal_z_score_1
z1_testG <- t(apply(test_count, 1, cal_z_score_1)) # scale counts to center
hmap1 <- subset(z1_testG, subset = rownames(z1_testG) %in% geneSet) # take out genes to analyze from zcounts

hmat1 <-  pheatmap(hmap1)

#scaling with cal_z_score_2
z2_testG <-t(scale(t(log2(1 + test_count)))) # scale counts to center
hmap2 <- subset(z2_testG, subset = rownames(z2_testG) %in% geneSet) # take out genes to analyze from zcounts

hmat2 <-  pheatmap(hmap2)

# scaling function 2 scales better on heatmap thus use this

```
  
###### To clean up data to generate heatmap (adding column annotations:
```{r}
conditions <- sGeorge[1, ]

sGeorge_clean <- sGeorge[-1 ,]
sGeorge_clean <- as.matrix(sGeorge_clean) # convert to matrix
mode(sGeorge_clean) = "numeric" # convert to numeric
counts <- sGeorge_clean
```
  
##### Writing the heatmap function
```{r}
library(pheatmap)
library(RColorBrewer)


# counts need to be numerical matrix

makeHeatmapGS <- function(counts, conditions, geneSet, title = "George Macrophage Polarization", 
                          showRow = T, c_rows = T, c_cols = T, sc = "none") {
  hcolour <- colorRampPalette(brewer.pal(8, "RdBu"))(100) # generate colour panel
  
  zcounts <- t(scale(t(log2(1 + counts))))  # scale counts to center
  hmap <- subset(zcounts, subset = rownames(zcounts) %in% geneSet) # take out genes to analyze from zcounts
  conditions <- as.character(conditions)
  col_cuts <- length(unique(conditions))
  ann_col <- data.frame("subtype" = conditions, row.names = colnames(counts)) 
  # row.names(ann_col) <- colnames(counts)
  # write heatmap parameters
  hmat <-  pheatmap(hmap, color=rev(hcolour), cluster_rows= c_rows, cluster_cols= c_cols, legend=T,
           show_rownames=showRow, annotation_col=ann_col, show_colnames=T,
           main=title, scale = sc,
           cutree_cols = col_cuts,
           breaks=seq(from=-max(abs(hmap)), to=max(abs(hmap)), length.out=100))
    return(hmat)
}

Macrophage_1 <- c("CD163", "CD63")
GMA <- makeHeatmapGS(counts = counts, conditions = conditions, geneSet = Macrophage_1)

setwd("/Users/benson/R_programming/RJourney")
pdf(file = "SCLC.George_macrophageCD163CD63.pdf", width = 15, height = 5)
grid::grid.newpage()
grid::grid.draw(GMA$gtable)
dev.off()
```
  
##### Testing out different genesets...
  
```{r}

M2genes <- c("MRC1", "CCL22", "FCER2", "ARG1", "CD163", "VEGF", "MSR1")

                
GMA <- makeHeatmapGS(counts = counts, conditions = conditions, geneSet = mGenes, c_rows = T,
                     sc = "none")
setwd("/Users/benson/R_programming/RJourney")
pdf(file = "SCLC.George_subtypes.pdf", width = 15, height = 5)
grid::grid.newpage()
grid::grid.draw(GMA$gtable)
dev.off()
```
  
##### Testing out the new markergene function  
```{r}

# heatmap needs to have marker genes for subtype added...

makeHeatmapMG <- function(counts, conditions, markerGenes, geneSet,
                        c_rows = T, c_cols = T, sc = "none",  
                        title = "George Macrophage Polarization", showRow = T) {
  hcolour <- colorRampPalette(brewer.pal(8, "RdBu"))(100) # generate colour panel
  zcounts <- t(scale(t(log2(1 + counts))))  # scale counts to center
  # make null matrices to subset genesets from markers and genes of interest  
  mg <- NULL
  gs <- NULL
  
  mg <- subset(zcounts, subset = rownames(zcounts) %in% markerGenes) # take out marker genes from zcounts
  gs <- subset(zcounts, subset = rownames(zcounts) %in% geneSet) # take out genes to analyze from zcounts
  mg <- as.matrix(mg)
  gs <- as.matrix(gs)
  hmap <- rbind(mg, gs)
  
  
  conditions <- as.character(conditions)
  col_cuts <- length(unique(conditions))
  row_gaps <- length(markerGenes)
  ann_col <- data.frame("subtype" = conditions, row.names = colnames(counts)) 
  # row.names(ann_col) <- colnames(counts)
  # write heatmap parameters
  
  hmat <-  pheatmap(hmap, color=rev(hcolour), cluster_rows=F, cluster_cols=T, legend=T,
           show_rownames=showRow, annotation_col=ann_col, show_colnames=T,
           main=title, 
           cutree_cols = col_cuts, gaps_row = row_gaps, 
           breaks=seq(from=-max(abs(hmap)), to=max(abs(hmap)), length.out=100))
    return(hmat)
}

mGenes = c("ASCL1", "YAP1", "NEUROD1", "POU2F3")
geneSet <- c("CD63", "CD163")

GMA <- makeHeatmapMG(counts = counts, conditions = conditions, markerGenes = mGenes, geneSet = macrophage)

pdf(file = "SCLC.George.4Subtypes_mall.pdf", width = 15, height = 5)
grid::grid.newpage()
grid::grid.draw(GMA$gtable)
dev.off()

# works well! now see if I can split further between M1 and M2 macrophages
```
  
###### Spliting clustering further between M1 and M2 macrophges

```{r}


makeHeatmap2GS <- function(counts, conditions, markerGenes, geneSet1, geneSet2,
                        title = "George Macrophage Polarization", showRow = T, showCol = T,
                        c_rows = T, c_cols = T, sc = "none") {
  hcolour <- colorRampPalette(brewer.pal(8, "RdBu"))(100) # generate colour panel
  zcounts <- t(scale(t(log2(1 + counts))))  # scale counts to center
  # make null matrices to subset genesets from markers and genes of interest  
  mg <- NULL
  gs <- NULL
  
  mg <- subset(zcounts, subset = rownames(zcounts) %in% markerGenes) # take out marker genes from zcounts
  gs1 <- subset(zcounts, subset = rownames(zcounts) %in% geneSet1) # take out genes to analyze from zcounts
  gs2 <- subset(zcounts, subset = rownames(zcounts) %in% geneSet2)
  mg <- as.matrix(mg)
  gs1 <- as.matrix(gs1)
  gs2 <- as.matrix(gs2)
  hmap <- rbind(mg, gs1, gs2)
  
  
  conditions <- as.character(conditions)
  col_cuts <- length(unique(conditions))
  row_gaps <- c(length(markerGenes), (length(markerGenes) + length(geneSet1)))
  ann_col <- data.frame("subtype" = conditions, row.names = colnames(counts)) 
  # row.names(ann_col) <- colnames(counts)
  # write heatmap parameters
  
  hmat <-  pheatmap(hmap, color=rev(hcolour), cluster_rows=c_rows, cluster_cols=c_cols, legend=T,
           show_rownames=showRow, annotation_col=ann_col, show_colnames=showCol,
           main= title, 
           cutree_cols = col_cuts, gaps_row = row_gaps, 
           breaks=seq(from=-max(abs(hmap)), to=max(abs(hmap)), length.out=100))
    return(hmat)
}

M1genes <- c("IL6", "IL1B", "TNF")
M2genes <- c("MRC1", "CCL22", "FCER2", "ARG1", "MSR1")

GMA <- makeHeatmap2GS(counts = counts, conditions = conditions, markerGenes = mGenes, geneSet1 = M1genes,
               geneSet2 = M2genes, title = "George Subtype M1 and M2 expression", c_rows = F, showCol = F,
               sc = "column")

pdf(file = "SCLC.George.4Subtypes_M1M2.pdf", width = 15, height = 5)
grid::grid.newpage()
grid::grid.draw(GMA$gtable)
dev.off()

```

###### data.frame():  
- creates data frames  
`data.frame(…, row.names = NULL, check.rows = FALSE, check.names = TRUE, fix.empty.names = TRUE, stringsAsFactors = default.stringsAsFactors())`
- ... arguments either form value or tag = value. Component names created based on tag
- row.names `NULL` or single integer or character string specifiying a column to be used as row names: gives row names of dataframe  
  
##### Conclusions:  
- from dataset, does not seem specific subtype associates with macrophage polarization, or if there is, it is difficult to discern due to low n of SCLC-P, SCLC-Y subtypes




a